### 2025-07-06 manuscript/part1.tex
expressions

\textbf{Type-3} grammars, or \emph{regular} grammars, generate exactly the class of regular languages. Their productions are restricted to the two equivalent styles:
\[
  right-reg.: \quad A \;\to\; aB \quad or \quad A \;\to\; a
  \qquad and \qquad
  left-reg.: \quad A \;\to\; Ba \quad or \quad A \;\to\; a,
\]
where \(A,B\) are nonterminals and \(a\) is a terminal. Regular expressions provide an alternative, declarative notation for these languages - each expression
can be mechanically transformed into a regular grammar, and vice versa. More broadly, every grammar in the Chomsky hierarchy admits an equivalent acceptor automaton;
in the case of Type-3 grammars, this correspondence yields deterministic or nondeterministic finite automata. Leveraging these conversions makes regular grammars
invaluable in practice (for example, in lexical analysis, pattern matching, and protocol verification), even though their expressive power is the most limited.
\textbf{Type-2} grammars, or \emph{context-free} grammars (CFGs), produce all context-free languages.
Their rules take the general form  \[A \;\to\; \alpha\] where \(A\) is a single nonterminal and \(\alpha\) is any string of terminals and nonterminals.
This additional flexibility captures nested, hierarchical structures-like balanced parentheses or most programming‐language syntaxes - that regular grammars cannot.
Each CFG is accepted by a corresponding pushdown automaton: the grammar expansions map naturally onto push and pop operations on the PDA’s stack, making the grammar-automaton
equivalence at this level another cornerstone of formal language theory.

>>>>

\textbf{Type-3} grammars, or \emph{regular} grammars, generate exactly the class of regular languages. Their productions are restricted to the two equivalent styles:
\[
  right-reg.: \quad A \;\to\; aB \quad or \quad A \;\to\; a
  \qquad and \qquad
  left-reg.: \quad A \;\to\; Ba \quad or \quad A \;\to\; a,
\]
where \(A,B\) are nonterminals and \(a\) is a terminal. Regular expressions provide an alternative, declarative notation for these languages - each expression
can be transformed into a regular grammar, and vice versa. Further, every grammar in the Chomsky hierarchy can be characterized by an equivalent acceptor automaton, which admits all input that ends in valid states as words of the language.
In the case of Type-3 grammars, this correspondence yields deterministic or nondeterministic finite automata. Leveraging these conversions makes regular grammars
invaluable in practice (for example, in lexical analysis, pattern matching, and protocol verification), even though their expressive power is the most limited.
\textbf{Type-2} grammars, or \emph{context-free} grammars (CFGs), produce all context-free languages.
Their rules take the general form  \[A \;\to\; \alpha\] where \(A\) is a single nonterminal and \(\alpha\) is any string of terminals and nonterminals.
This additional flexibility captures nested, hierarchical structures-like balanced parentheses or most programming-language syntaxes - that regular grammars cannot.
Each CFG is accepted by a corresponding pushdown automaton: the grammar expansions map naturally onto push and pop operations on the PDA's stack, making the grammar-automaton
equivalence at this level another cornerstone of formal language theory.

### 2025-07-06 manuscript/part2.tex
expressions

Most of the expression system is omitted here, as it closely follows the operator-precedence grammar pattern common to LL parsers.
Typically, expressions are divided into a hierarchy productions - such as Expression (handling addition and subtraction),
Term (for multiplication and division), Factor (for unary operators), and Primary (literals, identifiers, and parenthesized expressions)—and uses
“rest” productions (e.g. ExpressionRest → (“+” | “–”) Term ExpressionRest | ε) to eliminate left recursion while preserving left-associativity.
This arrangement guarantees that the parser can always choose the correct production with a single lookahead token, ensuring LL(1) compliance.
For a fully worked-out example, including support for exponentiation and custom binding powers, see the expression grammars in Langium’s samples repository on GitHub \cite{LangiumGit}

The lines depiceted here are chosen as they include a construct special? to MiniProb, the probabilistic assignment and the last level of the hierarchy.

@@

\subsection*{Types and terminals}
\begin{minted}[linenos, breaklines]{yaml}
Type:
    'bool' | IntType;
IntType:
    prefix=INT_PREFIX ({infer IntArray} '[' size=INT ']')?;
IntegerLiteral: // same name as infer IntLiteral creates minor conflicts
    sign=('+' | '-')? value=INT suffix=INT_PREFIX;

>>>>

Most of the expression system is omitted here, as it closely follows the \textit{operator-precedence grammar} pattern common to LL parsers \cite{operatorPrecedence}.
Typically, a grammar defines a series of nonterminals - each corresponding to a distinct operator-precedence level (from primaries up to full expressions) - and structures its productions so as to eliminate left recursion while preserving left-associative evaluation at every level.
This arrangement guarantees that the parser can always choose the correct production with a single lookahead token, ensuring LL(1) compliance.

\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  \begin{forest}
    for tree={
      grow=south,
      edge={-stealth},
      l sep=0.2em,
      align=right,
    }
    [Expression
      [ProbabilisticAssignment?
        [LogicalOr\,(\texttt{||})
          [LogicalAnd\,(\texttt{\&\&})
            [Comparison\,(\texttt{< <= > >= == !=})
              [Term\,(\texttt{+ -})
                [Factor\,(\texttt{*})
                  [Division\,(\texttt{/ \%})
                    [Unary\,(\texttt{!})
                      [Primary]
                    ]
                  ]
                ]
              ]
            ]
          ]
        ]
      ]
    ]
  \end{forest}
  \caption{Operator‐precedence hierarchy}
  \label{fig:op-prec-tree}
\end{wrapfigure}

In Langium, rules can be made to share other base classes than \code{AstNode} by utilizing the \code{infers} keyword. This is especially useful for creating a layered expression system, as all the levels infer from the top-most expression (line 3), which follows logical abstraction and enables dynamic and easy resolution of parsed expressions.
The lines depicted here have been selected because they include the \code{Probabilistic Assignment}, a construct particular to MiniProb (see \ref{sec:miniprob}), alongside the \code{Primary} rule, which defines the core forms to which all expressions ultimately resolve.
The topmost level of the precedence hierarchy was originally occupied by the disjunction operator (\code{||}), but was ultimately edged out by the probabilistic assignment. This behavior is enabled by another Langium specific command \code{infers}, which instructs a rule to create a new AST node based on the structure of the input.
In this case, an \code{Expression} is by default parsed as a \code{LogicalOr}, but immediately becomes a \code{Probabilistic Assignment} node as soon as the first curly brace is recognized (lines 2, 4). The remainder of the operator precedence follows conventional conventions, with the exception that \code{*} binds more loosely than division and modulo (see Figure~\ref{fig:op-prec-tree}).

At the base of the precedence hierarchy is the \code{Primary} rule, to which every expression ultimately reduces. A \code{Primary} can be either a boolean literal, an integer literal, or a variable reference (\code{Lval}); treating \code{Lval} as a \code{Primary} makes variable accesses first-class expressions,
as is conventional in many languages. In addition, the production allows any expression to be wrapped in arbitrarily many parentheses which only delegate to the inner expression without altering the meaning.

@@

\subsection*{Types and terminals}

\begin{minted}[linenos, breaklines]{yaml}
Type:
    'bool' | IntType;
IntType:
    prefix=INT_PREFIX ({infer IntArray} '[' size=INT ']')?;
IntegerLiteral: // same name as infer IntLiteral creates minor conflicts
    sign=('+' | '-')? value=INT suffix=INT_PREFIX;
